From Reversible Programs to \\ Univalent Universes and Back

### 2021-10-27 Pi/paper/ufin.tex
Restore some HoTT text from appendix
Fixes #141

We use the type theory of the HoTT book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}, that is, we use
intensional Martin-L\"{o}f Type Theory, with a (univalent) universe $\UU$, and a few Higher Inductive Types (HITs) for
propositional and set truncation, and set-quotients. All arguments will hold in a Cubical Type
Theory~\cite*{cohenCubicalTypeTheory2018,angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}
as well.  In~\cref{app:identitytypes},~\cref{app:homotopytypes}, and~\cref{app:hits}, we review the basics of identity
types, homotopy types, and HITs and refer the reader to the book for more details.

@@

\begin{toappendix}
  \subsection{Identity Types}
  \label{app:identitytypes}

  Given two terms $x:A$ and $y:A$, we write $x \id_{A} y$, or simply $x \id y$, for the identity type, which is the type
  of equalities or identifications between them. The identity type is generated by reflexivity $\refl_{x} : x \id_{A} x$,
  and the eliminator for the identity type is given by path induction or the $J$-rule (\cref{def:path-induction}). This
  construction can be iterated, giving the identity type between two terms of an identity type, repeating ad infinitum.
  Using the iterated identity type for morphisms, each type is equipped with the structure of a weak $\infty$-groupoid,
  where each morphism satisfies groupoid laws only upto a higher one. Given an arbitrary type (or groupoid) $A$, we list
  some laws that are provable using path induction.

@@

  \begin{gather*}
    \begin{aligned}
      \term{\inv{\blank}}      & : (x \id_{A} y) \to (y \id_{A} x)                   \\
      \term{\blank\comp\blank} & : (x \id_{A} y) \to (y \id_{A} z) \to (x \id_{A} z)
    \end{aligned}
    \qquad
    \begin{aligned}
      \term{assoc} & : (p : x \id_{A} y)  (q : y \id_{A} z) (r : z \id_{A} w) \\
                   & \to (p \comp q) \comp r \id p \comp (q \comp r)          \\
      \term{invr}  & : (p : x \id_{A} y) \to p \comp \inv{p} \id \refl_{x}
    \end{aligned}
  \end{gather*}

  \noindent A homotopy between functions $f \htpy g$ is given by pointwise equality between them $\dfun{x:A}{f(x) \id_{B} g(x)}$.
  The identity type for functions is equivalent to homotopies between them ${(f \id_{A \to B} g)} \eqv {(f \htpy g)}$, by
  function extensionality. An equivalence between types $A \eqv B$ is given by a pair of functions between them which
  compose to the identity, $f \comp g \htpy \idfunc_{B}$ and $g \comp f \htpy \idfunc_{A}$, and this is equivalent to the
  identity type for the universe, $(A \id_{\UU} B) \eqv (A \eqv B)$, by univalence.

  Functions between types are functors between groupoids. Given a function $f : A \to B$, there is a functorial action
  on the paths given by $\term{ap}$. Type families, that is, types indexed by terms, are simply functions from a type to
  the universe, such as $A \to \UU$. For a type family $P : A \to \UU$ and a point $x : A$, the type $P(x)$ is called
  the fiber over $x$.  The $\term{transport}$ operation, named $\term{transport}/term{tr}$, lifts paths in the indexing
  type to functions between fibers.
  %which is an $A$-indexed family of groupoids.

  \begin{gather*}
    \term{ap}_{f} : \dfun{x,y:A}{x \id_{A} y \to f(x) \id_{B} f(y)}
    \qquad
    \term{transport}_{P} : \dfun{x,y:A}{x \id_{A} y \to P(x) \to P(y)}
  \end{gather*}

  The type $\dsum*{x:A}{P(x)}$ is the collection of all the fibers and is called the total space of $P$. The first
  projection ${\pi_1 : \dsum*{x:A}{P(x)} \to A}$ from the total space to the base space $A$ has the structure of a
  fibration, that is, there is a lifting operation (\cref{fig:lift}) which lifts paths in the base space to paths in the
  total space. Given a path $p : x \id_{A} y$ in the base space, and $u : P(x)$ a point in the fiber over $x$, we have:
  \[
    \term{lift}(u,p) : (x , u) \id_{\dsum{x:A}{P(x)}} (y , \transport{P}{p}{u})
  \]

  % where $\tr{p}{u}$ is shorthand for $$.

  \begin{figure}
    \begin{center}
      \begin{tikzpicture}[yscale=.5,xscale=2]
        \draw (0,0) arc (-90:170:8ex) node[anchor=south east] {$A$} arc (170:270:8ex);
        \draw (0,6) arc (-90:170:10ex) node[anchor=south east] {$\dsum{x:A}{P(x)}$} arc (170:270:10ex);
        \draw[->] (0,5.8) -- node[auto] {$\fst$} (0,3.2);
        \node[circle,fill,inner sep=1pt,label=left:{$x$}] (b1) at (-.5,1.4) {};
        \node[circle,fill,inner sep=1pt,label=right:{$y$}] (b2) at (.5,1.4) {};
        \draw[decorate,decoration={snake,amplitude=1}] (b1) -- node[auto,swap] {$p$} (b2);
        \node[circle,fill,inner sep=1pt,label=left:{$\pair{x,u}$}] (b1) at (-.5,7.2) {};
        \node[circle,fill,inner sep=1pt,label=right:{$\pair{y,\transport*{P}{p}{u}}$}] (b2) at (.5,7.2) {};
        \draw[decorate,decoration={snake,amplitude=1}] (b1) -- node[auto] {$\term{lift}(u,p)$} (b2);
      \end{tikzpicture}
    \end{center}
    \caption{Lifting operation in $P$}
    \label{fig:lift}
  \end{figure}

@@

Functions between types are functors between groupoids, and type families (or functions to the universe) are indexed
families of groupoids. A type family $P : A \to \UU$ comes equipped with a functor $\pi_1 : \dsum*{x:A}{P(x)} \to A$
which has a lifting operation giving it the structure of a fibration. The $\term{transport}$ operation lifts paths in
the base space to functions between the fibers.  Using the groupoid structure of $A$, for any $x, y : A$ and a path 
$p : x \id_{A} y$, $\term{transport}(p)$ and $\term{transport}(\inv{p})$ form an equivalence.
% We name this map $\tptEqv{P}$ which lifts paths to equivalences of fibers.
\[
  \tptEqv{P} : (x \id_{A} y) \to (P(x) \eqv P(y))
\]

\noindent The type families (or fibrations) we are interested in are the ones where paths in the base space completely determine
the equivalences in the fibers -- these are called univalent
fibrations~\cite*{kapulkinUnivalenceSimplicialSets2018,kapulkinSimplicialModelUnivalent2021,christensenCharacterizationUnivalentFibrations2015}.

@@

Univalent fibrations were introduced by~\citet*{kapulkinUnivalenceSimplicialSets2018}, to build a model of Voevodsky's
\emph{univalence} principle in simplicial sets. Indeed, univalence characterises paths in the universe as equivalences
between types, which follows from the canonical fibration $\idfunc : \UU \to \UU$ being univalent.

@@

  A type is \emph{contractible} (-2-type) if it has a unique element, that is, there is a center of contraction and every
  other point is equal to it. A type is a \emph{proposition} (-1-type) if its equality types are contractible, that is, it
  has at most one inhabitant. Iterating this, we can define sets or 0-types (whose equality types are propositions) and
  1-groupoids or 1-types (whose equality types are sets), and similarly, higher homotopy $n$-types.

  \begin{gather*}
    \begin{aligned}
      \isContr{A} & \defeq \dsum{x:A}{\dfun{y:A}{y \id x}} \\
      \isProp{A}  & \defeq \dfun{x,y:A}{\isContr{x \id y}}
    \end{aligned}
    \qquad
    \begin{aligned}
      \isSet{A} & \defeq \dfun{x,y:A}{\isProp{x \id y}} \\
      \isGpd{A} & \defeq \dfun{x,y:A}{\isSet{x \id y}}
    \end{aligned}
  \end{gather*}

@@

  \begin{definition}[Propositional Truncation]
    \label{def:prop-trunc}
    Given a type $A$, the propositional truncation $\Trunc[-1]{A}$, or simply $\Trunc{A}$, is a higher inductive type
    generated by the following constructors,
    \begin{itemize}
      \item an inclusion function $\trunc{\blank} : A \to \Trunc{A}$,
      \item for each $x, y : \Trunc{A}$, a path $\term{trunc}(x,y) : x \id_{\Trunc{A}} y$,
    \end{itemize}
    such that, given any type $B$ with
    \begin{itemize}
      \item a function $g : A \to B$,
      \item for each $x, y : B$, a path $\term{trunc*}(x,y) : x \id_{B} y$,
    \end{itemize}
    there is a unique function $f : \Trunc{A} \to B$ such that,
    \begin{itemize}
      \item $f(\trunc{a}) \equiv g(a)$
      \item for each $x, y : \Trunc{A}$, $\ap{f}{\term{trunc}(x,y)} \id_{B} \term{trunc*}(f(x),f(y))$.
    \end{itemize}
  \end{definition}

  \begin{definition}[${\fib_{f}} : B \to \UU$]
    \label{def:fib}
    The fiber of $f : A \to B$ at $b : B$ is
    \[
      \fib_{f}(b) \defeq \dsum{a:A}{f(a) \id_{B} b}.
    \]
  \end{definition}

  \begin{definition}[${\term{im}} : (f : A \to B) \to \UU$]
    \label{def:im}
    The image of $f$ is the (-1)-truncation of its fiber.
    \[
      \im{f} \defeq \dsum{b:B}{\Trunc[-1]{\fib_{f}(b)}}
    \]
  \end{definition}

  \begin{lemma}
    The following are equivalent.
    \begin{enumerate}
      \item $f : A \to B$ is an equivalence.
      \item $f$ has a left and right inverse.
      \item $f$ has contractible fibers.
    \end{enumerate}
  \end{lemma}

  Another HIT that we use is the set-quotient $\quot{A}{R}$ which takes an set $A$ and a relation $R : A \to A \to \UU$.
  It has an inclusion of points $\quotinc : A \to \quot{A}{R}$, and adds paths between related pairs of elements
  $\quotrel : R(x,y) \to \quotinc(x) \id_{\quot{A}{R}} \quotinc(y)$ (see~\cref{def:set-quot}).  We recall that the
  quotient is \emph{effective} if $R$ is a prop-valued equivalence relation, that is, $R(x,y)$ holds iff $(q(x)
  \id_{\quot{A}{R}} q(y))$.

  \begin{definition}[Set Quotient]
    \label{def:set-quot}
    Given a type $A$ which is a set, and a relation $R : A \to A \to \UU$, the set-quotient $\quot{A}{R}$ is the higher
    inductive type generated by
    \begin{itemize}
      \item an inclusion function $\quotinc : A \to \quot{A}{R}$,
      \item for each $x, y : A$ such that $R(x,y)$, a path $\quotinc(x) \id_{\quot{A}{R}} \quotinc(y)$,
      \item a set truncation, for each $x, y : \quot{A}{R}$ and $r, s : x \id_{\quot{A}{R}} y$, we have $r \id s$,
    \end{itemize}
    with an appropriate induction principle.
  \end{definition}

@@

\begin{propositionrep}[Univalent Subuniverse]
  \label{prop:univsub}
  A universe predicate is a type family $P : \UU \to \UU$ whose fibers are propositions, that is, $P(X)$ is a
  proposition for every $X$. Given such a predicate $P$, the fibration ${\fst : \dsum*{X:\UU}{P(X)} \to \UU}$ is
  univalent and generates a univalent subuniverse ${\UU_{P} \defeq (\dsum*{X:\UU}{P(X)}, \fst)}$.~\footnote{Univalent
    typoids~\cite{petrakisUnivalentTypoids2019a} are a different presentation of univalent subuniverses.}
\end{propositionrep}

>>>>

We use the type theory of the HoTT book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}, that is, we use
intensional Martin-L\"{o}f Type Theory, with a (univalent) universe $\UU$, and a few Higher Inductive Types (HITs) for
propositional and set truncation, and set-quotients. All arguments will hold in a Cubical Type
Theory~\cite*{cohenCubicalTypeTheory2018,angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}
as well. We review the basics of identity types, homotopy types, and some HITs that we use, and refer the reader to the
book for more details.

@@

\subsection{Identity Types}
\label{app:identitytypes}

Given two terms $x:A$ and $y:A$, we write $x \id_{A} y$, or simply $x \id y$, for the identity type, which is the type
of equalities or identifications between them. The identity type is generated by reflexivity $\refl_{x} : x \id_{A}
  x$, and the eliminator for the identity type is given by path induction or the $J$-rule (see~\cref{def:path-induction}
in the appendix). This construction can be iterated, giving the identity type between two terms of an identity type,
repeating ad infinitum. Using the iterated identity type for morphisms, each type is equipped with the structure of a
weak $\infty$-groupoid, where each morphism satisfies groupoid laws only upto a higher one.

\begin{toappendix}
  Given an arbitrary type (or groupoid) $A$, we list some laws that are provable using path induction.

@@

  \begin{gather*}
    \begin{aligned}
      \term{\inv{\blank}}      & : (x \id_{A} y) \to (y \id_{A} x)                   \\
      \term{\blank\comp\blank} & : (x \id_{A} y) \to (y \id_{A} z) \to (x \id_{A} z)
    \end{aligned}
    \qquad
    \begin{aligned}
      \term{assoc} & : (p : x \id_{A} y)  (q : y \id_{A} z) (r : z \id_{A} w) \\
                   & \to (p \comp q) \comp r \id p \comp (q \comp r)          \\
      \term{invr}  & : (p : x \id_{A} y) \to p \comp \inv{p} \id \refl_{x}
    \end{aligned}
  \end{gather*}
\end{toappendix}

A homotopy between functions $f, g : A \to B$, written $f \htpy g$, is given by pointwise equality between them,
$\dfun*{x:A}{f(x) \id_{B} g(x)}$. The identity type for functions is equivalent to homotopies between them ${(f \id_{A
        \to B} g)} \eqv {(f \htpy g)}$, by $\term{funext}$ and $\term{happly}$. An equivalence between types $A \eqv B$ is given
by a pair of functions between them which compose to the identity, $f \comp g \htpy \idfunc_{B}$ and $g \comp f \htpy
  \idfunc_{A}$ (see~\cref{prop:three-chars-equiv} in the appendix), and this is equivalent to the identity type for the
universe, $(A \id_{\UU} B) \eqv (A \eqv B)$, by univalence.

Functions between types are functors between groupoids. Given a function $f : A \to B$, there is a functorial action on
the paths given by $\term{ap}$. Type families, that is, types indexed by terms, are simply functions from a type to the
universe, such as $A \to \UU$, which is an $A$-indexed family of groupoids. For a type family $P : A \to \UU$ and a
point $x : A$, the type $P(x)$ is the fiber over $x$.  The $\term{transport}$ operation, named
$\term{transport}/\term{tr}$, lifts paths in the indexing type to functions between fibers.
\begin{gather*}
  \term{ap}_{f} : \dfun*{x,y:A}{x \id_{A} y \to f(x) \id_{B} f(y)}
  \qquad
  \term{transport}_{P} : \dfun*{x,y:A}{x \id_{A} y \to P(x) \to P(y)}
\end{gather*}

The type $\dsum*{x:A}{P(x)}$ is the collection of all the fibers and is the total space of $P$. The first projection
${\pi_1 : \dsum*{x:A}{P(x)} \to A}$ from the total space to the base space $A$ has the structure of a fibration, that
is, there is a lifting operation (\cref{fig:lift} in the appendix) which lifts paths in the base space to paths in the
total space. Given a path $p : x \id_{A} y$ in the base space, and $u : P(x)$ a point in the fiber over $x$, we have:
\[
  \term{lift}(u,p) : (x , u) \id_{\dsum*{x:A}{P(x)}} (y , \transport{P}{p}{u})
\]

% where $\tr{p}{u}$ is shorthand for $$.
\begin{toappendix}
  \begin{figure}
    \begin{center}
      \begin{tikzpicture}[yscale=.5,xscale=2]
        \draw (0,0) arc (-90:170:8ex) node[anchor=south east] {$A$} arc (170:270:8ex);
        \draw (0,6) arc (-90:170:10ex) node[anchor=south east] {$\dsum{x:A}{P(x)}$} arc (170:270:10ex);
        \draw[->] (0,5.8) -- node[auto] {$\fst$} (0,3.2);
        \node[circle,fill,inner sep=1pt,label=left:{$x$}] (b1) at (-.5,1.4) {};
        \node[circle,fill,inner sep=1pt,label=right:{$y$}] (b2) at (.5,1.4) {};
        \draw[decorate,decoration={snake,amplitude=1}] (b1) -- node[auto,swap] {$p$} (b2);
        \node[circle,fill,inner sep=1pt,label=left:{$\pair{x,u}$}] (b1) at (-.5,7.2) {};
        \node[circle,fill,inner sep=1pt,label=right:{$\pair{y,\transport*{P}{p}{u}}$}] (b2) at (.5,7.2) {};
        \draw[decorate,decoration={snake,amplitude=1}] (b1) -- node[auto] {$\term{lift}(u,p)$} (b2);
      \end{tikzpicture}
    \end{center}
    \caption{Lifting operation in $P$}
    \label{fig:lift}
  \end{figure}
\end{toappendix}

@@

Using the groupoid structure of $A$, for any $x, y : A$ and a path $p : x \id_{A} y$, $\term{transport}(p)$ and
$\term{transport}(\inv{p})$ form an equivalence.
\[
  \tptEqv{P} : (x \id_{A} y) \to (P(x) \eqv P(y))
\]

\noindent The type families (or fibrations) we are interested in are the ones where paths in the base space completely
determine the equivalences in the fibers -- these are called univalent
fibrations~\cite*{kapulkinUnivalenceSimplicialSets2012,kapulkinSimplicialModelUnivalent2021,christensenCharacterizationUnivalentFibrations2015}.

@@

Univalent fibrations were introduced by~\citet*{kapulkinUnivalenceSimplicialSets2012}, to build a model of Voevodsky's
\emph{univalence} principle in simplicial sets. Indeed, univalence characterises paths in the universe as equivalences
between types, which follows from the canonical fibration $\idfunc : \UU \to \UU$ being univalent.

@@

A type is \emph{contractible} (-2-type) if it has a unique element, that is, there is a center of contraction and every
other point is equal to it. A type is a \emph{proposition} (-1-type) if its equality types are contractible, that is, it
has at most one inhabitant. Iterating this, we can define sets or 0-types (whose equality types are propositions) and
1-groupoids or 1-types (whose equality types are sets), and similarly, higher homotopy $n$-types.
\begin{gather*}
  \begin{aligned}
    \isContr{A} & \defeq \dsum*{x:A}{\dfun*{y:A}{y \id x}} \\
    \isProp{A}  & \defeq \dfun*{x,y:A}{\isContr{x \id y}}
  \end{aligned}
  \qquad
  \begin{aligned}
    \isSet{A} & \defeq \dfun*{x,y:A}{\isProp{x \id y}} \\
    \isGpd{A} & \defeq \dfun*{x,y:A}{\isSet{x \id y}}
  \end{aligned}
\end{gather*}

@@

\begin{toappendix}
  \begin{definition}[Propositional Truncation]
    \label{def:prop-trunc}
    Given a type $A$, the propositional truncation $\Trunc[-1]{A}$, or simply $\Trunc{A}$, is a higher inductive type
    generated by the following constructors,
    \begin{itemize}
      \item a function $\trunc{\blank} : A \to \Trunc{A}$,
      \item for each $x, y : \Trunc{A}$, a path $\term{trunc}(x,y) : x \id_{\Trunc{A}} y$,
    \end{itemize}
    such that, given any type $B$ with
    \begin{itemize}
      \item a function $g : A \to B$,
      \item for each $x, y : B$, a path $\term{trunc*}(x,y) : x \id_{B} y$,
    \end{itemize}
    there is a unique function $f : \Trunc{A} \to B$ such that,
    \begin{itemize}
      \item $f(\trunc{a}) \equiv g(a)$
      \item for each $x, y : \Trunc{A}$, $\ap{f}{\term{trunc}(x,y)} \id_{B} \term{trunc*}(f(x),f(y))$.
    \end{itemize}
  \end{definition}
\end{toappendix}

\begin{toappendix}
  \begin{definition}[${\fib_{f}} : B \to \UU$]
    \label{def:fib}
    The fiber of $f : A \to B$ at $b : B$ is
    \[
      \fib_{f}(b) \defeq \dsum*{a:A}{f(a) \id_{B} b}.
    \]
  \end{definition}

  \begin{definition}[${\term{im}} : (f : A \to B) \to \UU$]
    \label{def:im}
    The image of $f$ is the (-1)-truncation of its fiber.
    \[
      \im{f} \defeq \dsum*{b:B}{\Trunc[-1]{\fib_{f}(b)}}
    \]
  \end{definition}

  \begin{proposition}
    \label{prop:three-chars-equiv}
    The following are equivalent.
    \begin{enumerate}
      \item $f : A \to B$ is an equivalence.
      \item $f$ has a left and right inverse.
      \item $f$ has contractible fibers.
    \end{enumerate}
  \end{proposition}
\end{toappendix}

Another HIT that we use is the set-quotient $\quot{A}{R}$ which takes a set $A$ and a binary relation $R : {A \to A \to
\UU}$. It has a mapping of points $\quotinc : A \to \quot{A}{R}$, and a constructor that adds paths between related
pairs of elements $\quotrel : R(x,y) \to \quotinc(x) \id_{\quot{A}{R}} \quotinc(y)$ (see~\cref{def:set-quot}).  We
recall that the quotient is \emph{effective} if $R$ is a prop-valued equivalence relation, that is, $R(x,y)$ holds iff
$(q(x) \id_{\quot{A}{R}} q(y))$.

\begin{toappendix}
  \begin{definition}[Set Quotient]
    \label{def:set-quot}
    Given a type $A$ which is a set, and a relation $R : A \to A \to \UU$, the set-quotient $\quot{A}{R}$ is the higher
    inductive type generated by
    \begin{itemize}
      \item an inclusion function $\quotinc : A \to \quot{A}{R}$,
      \item for each $x, y : A$ such that $R(x,y)$, a path $\quotinc(x) \id_{\quot{A}{R}} \quotinc(y)$,
      \item a set truncation, for each $x, y : \quot{A}{R}$ and $r, s : x \id_{\quot{A}{R}} y$, we have $r \id s$,
    \end{itemize}
    with an appropriate induction principle.
  \end{definition}
\end{toappendix}

@@

\begin{propositionrep}[Univalent Subuniverse]
  \label{prop:univsub}
  A universe predicate is a type family $P : \UU \to \UU$ whose fibers are propositions, that is, $P(X)$ is a
  proposition for every $X$. Given such a predicate $P$, the fibration ${\fst : \dsum*{X:\UU}{P(X)} \to \UU}$ is
  univalent and generates a univalent subuniverse ${\UU_{P} \defeq (\dsum*{X:\UU}{P(X)}, \fst)}$.~\footnote{Univalent
  typoids~\cite{petrakisUnivalentTypoids2019} are a different presentation of univalent subuniverses.}
\end{propositionrep}
